---
title: "An exploration of the gapminder dataframe"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In this document, we will be exploring the `gapminder` dataset. To load the `gapminder` dataset, we write
```{r}
library(gapminder)
```
## Exploring the gapminder dataset

The function `summary` provides some basic analysis of the dataset
```{r gapminder}
summary(gapminder)
```
We can see that `gapminder` provides data on population, gpd, and life expectancy for a number of countries, recorded over a number of years from 1952-2007. Among all countries and all years, the minimum life expectancy was 23.60 years, while the maximum life expectancy was 82.60. We can also obtain this information by the `max` and `min` functions on the `lifeEXP` column:
```{r}
min(gapminder$lifeExp)
max(gapminder$lifeExp)
```
Also useful is the `str` function.
```{r}
str(gapminder)
```
We see that data was recorded from 142 countries, every 5 years.

Suppose we want to find which country, and in what year, the minimum/maximum life expectancy above was. We can use the `which` function to achieve this. 
```{r}
min_country = gapminder[which.min(gapminder$lifeExp),]
min_country
max_country = gapminder[which.max(gapminder$lifeExp),]
max_country
```

Our conclusion is that Rwanda in 1992 had the lowest life expectancy, while Japan in 2007 had the highest life expectancy.

## Including Plots

Let us visualise how the life expectancy of Rwanda and Japan, respectively, evolves over the years sampled in the data. The relevant data `year` and `lifeEXP` are columns 3 and 4 of `gapminder`.
```{r}
Rwanda_data_location = which(gapminder$country == "Rwanda")
Japan_data_location = which(gapminder$country == "Japan")
Rwanda = gapminder[Rwanda_data_location, 3:4]
Japan = gapminder[Japan_data_location, 3:4]
```
Now we plot.
```{r}
plot(Rwanda, main = "Rwanda")
plot(Japan, main  = "Japan")
```

We observe that the long term trend is an increase in life expectancy for both countries, however an event in Rwanda in the early 1990's caused a sudden drop in life expectancy there (this event being well-known in the literature).
